---
title: "diff_accessibility"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
dyn.load("/home/rstudio/libs/libxml2.so.2")
library(DESeq2)
library(dplyr)
library(stringr)
library(apeglm)
library(ggplot2)
dyn.load("/home/rstudio/libs/libgsl.so.25")
dyn.load("/home/rstudio/libs/libgslcblas.so.0")
library(TFBSTools)
library(Biostrings)
```

Set significance parameters:

```{r, include=T}
# Joaquina's parameters
# fdr = 0.01
# min.l2fc = 2 # Min log2(fold change)
# min.baseMean = 100

# fdr = 0.01
# min.l2fc = 2 # Min log2(fold change)
# min.baseMean = 80

# fdr = 0.01
# min.l2fc = 2 # Min log2(fold change)
# min.baseMean = 60

# fdr = 0.01
# min.l2fc = 2 # Min log2(fold change)
# min.baseMean = 40

# fdr = 0.01
# min.l2fc = 2 # Min log2(fold change)
# min.baseMean = 20

# fdr = 0.01
# min.l2fc = 1.5 # Min log2(fold change)
# min.baseMean = 20

# fdr = 0.01
# min.l2fc = 1 # Min log2(fold change)
# min.baseMean = 20

# fdr = 0.01
# min.l2fc = 0.5 # Min log2(fold change)
# min.baseMean = 20

# fdr = 0.02
# min.l2fc = 0.5 # Min log2(fold change)
# min.baseMean = 20

# fdr = 0.03
# min.l2fc = 0.5 # Min log2(fold change)
# min.baseMean = 20

# fdr = 0.04
# min.l2fc = 0.5 # Min log2(fold change)
# min.baseMean = 20

fdr = 0.05
min.l2fc = 0.5 # Min log2(fold change)
min.baseMean = 20

# ---

# fdr = 0.01
# min.l2fc = 2 # Min log2(fold change)
# min.baseMean = 0

# fdr = 0.01
# min.l2fc = 0 # Min log2(fold change)
# min.baseMean = 0

# fdr = 0.05
# min.l2fc = 0 # Min log2(fold change)
# min.baseMean = 0
```

Prepare the general function for the analysis of differential accessibility:

```{r, include=T}
analyse_diff_accessibility = function(cond.target.long, cond.ref.long, 
                                      cond.target.short, cond.ref.short) {
  # Find differentially accessible regions
  p = read.delim(paste0("../results/select_samples/featureCounts_",  
                 cond.ref.long, "_vs_", cond.target.long, ".tsv")) %>%
  dplyr::select(-X)

  names(p) = gsub(pattern = ".mLb.clN.bam", replacement = "", x = names(p))
  
  p.ann = p[, c(1:6)]
  
  p.clean = p[, !(names(p) %in% names(p)[1:6])]
  
  rownames(p.clean) = p.ann$Geneid
  
  p.cond = data.frame(condition = stringr::str_extract(names(p.clean), 
                                                       paste0("(", cond.target.short, "|", cond.ref.short, ")")))
  
  row.names(p.cond) = names(p.clean)
  
  p.cond$condition = factor(p.cond$condition, levels = c(cond.target.short, cond.ref.short))
  
  p.dds = DESeqDataSetFromMatrix(countData = p.clean,
                                 colData = p.cond,
                                 design = ~ condition)
  p.dds = DESeq(p.dds)
  
  p.res = results(p.dds, contrast = c("condition", cond.target.short, cond.ref.short), alpha = fdr)
  
  # Print out stats
  cat(cond.target.long, "vs", cond.ref.long, "\n")
  
  cat("Number of significantly differentially accessible elements (FDR =", round(fdr * 100, 0), "%):", 
    sum(p.res$padj < fdr & abs(p.res$log2FoldChange) > min.l2fc & p.res$baseMean > min.baseMean, na.rm = T), 
    "\n")

  p.n.up = nrow(as.data.frame(p.res) %>% 
    filter(padj < fdr & abs(log2FoldChange) > min.l2fc & baseMean > min.baseMean) %>%
    filter(log2FoldChange > 0))
  
  cat("Number of upregulated elements (FDR =", round(fdr * 100, 0), "%):", 
      p.n.up,
      "\n")
  
  p.n.down = nrow(as.data.frame(p.res) %>% 
    filter(padj < fdr & abs(log2FoldChange) > min.l2fc & baseMean > min.baseMean) %>%
    filter(log2FoldChange < 0))
  
  cat("Number of downregulated elements (FDR =", round(fdr * 100, 0), "%):", 
      p.n.down,
      "\n")
  
  p.n.all = nrow(as.data.frame(p.res))
  
  cat("The total number of elements:",
      p.n.all,
      "\n")
  
  cat("The percentage of significantly differentially accessible elements:",
      round((p.n.up + p.n.down) / p.n.all * 100, 2))
  
  cat("\n")
  
  # Do PCA
  p.rld = rlog(p.dds)
  
  # Based on the code from DESeq2 vignette:
  # https://bioconductor.org/packages/devel/bioc/vignettes/DESeq2/inst/doc/DESeq2.html#principal-component-plot-of-the-samples
  p.pcaData = plotPCA(p.rld, returnData = T)
  
  p.percentVar = round(100 * attr(p.pcaData, "percentVar"), 2)
  
  pca.plot = ggplot(p.pcaData, aes(PC1, PC2, color = condition)) +
    geom_point(size = 2) +
    xlab(paste0("PC1: ", p.percentVar[1], "% variance")) +
    ylab(paste0("PC2: ", p.percentVar[2], "% variance")) + 
    theme_classic()
  
  ggsave(filename = paste0("../r_results/diff_accessibility/plots/pca_", cond.target.long, "_vs_", cond.ref.long,
                           "_fdr", fdr, "_min-l2fc", min.l2fc, "_min-baseMean", min.baseMean, ".pdf"),
         plot = pca.plot)
  
  # Plot the proportions of differentially regulated regions
  p.freq.up = p.n.up / p.n.all

  p.freq.down = p.n.down / p.n.all
  
  p.freq.nonsign = 1 - p.freq.up - p.freq.down
  
  p.freq.df = data.frame(freq     = c(p.freq.up, p.freq.down, p.freq.nonsign),
                         category = c("Up",      "Down",      "Non-significant"),
                         element_type = "Elements") %>%
    dplyr::mutate(category = factor(category, levels = category))
  
  proportions.plot = ggplot(p.freq.df) +
    geom_col(aes(x = element_type,
                 y = freq,
                 fill = category)) + 
    theme_classic()
  
  ggsave(filename = paste0("../r_results/diff_accessibility/plots/deg-proportions_", cond.target.long, "_vs_", cond.ref.long, 
                           "_fdr", fdr, "_min-l2fc", min.l2fc, "_min-baseMean", min.baseMean, ".pdf"),
         plot = proportions.plot)
  
  # Save the result DESeq table
  saveRDS(p.res, file = paste0("../r_results/diff_accessibility/tables/deseq2-results_", cond.target.long, "_vs_", cond.ref.long, 
                               "_fdr", fdr, "_min-l2fc", min.l2fc, "_min-baseMean", min.baseMean, ".rds"))
  
  # Save the result DESeq object
  saveRDS(p.dds, file = paste0("../r_results/diff_accessibility/tables/deseq2-dds_", cond.target.long, "_vs_", cond.ref.long, 
                               "_fdr", fdr, "_min-l2fc", min.l2fc, "_min-baseMean", min.baseMean, ".rds"))
  
  return(p.res)
}
```

## Find regions differentially regulated in WT-D11 vs WT-D7

```{r, include=T}
# p1
p1.11.7.res = analyse_diff_accessibility(cond.target.long =  "WT_D11_p1",
                                         cond.ref.long =     "WT_D7_p1",
                                         cond.target.short = "D11",
                                         cond.ref.short =    "D7")
# p2
p2.11.7.res = analyse_diff_accessibility(cond.target.long =  "WT_D11_p2",
                                         cond.ref.long =     "WT_D7_p2",
                                         cond.target.short = "D11",
                                         cond.ref.short =    "D7")
# pMN
pM.11.7.res = analyse_diff_accessibility(cond.target.long =  "WT_D11_pM",
                                         cond.ref.long =     "WT_D7_pM",
                                         cond.target.short = "D11",
                                         cond.ref.short =    "D7")
```

Upregulated regions outnumber downregulated regions by 2-3 times, which is expected if NFIA massively opens up chromatin.

## Find regions differentially regulated in MUT-D11 vs WT-D11

```{r, include=T}
# p1
p1.mut.wt.res = analyse_diff_accessibility(cond.target.long  = "MUT_D11_p1",
                                           cond.ref.long     = "WT_D11_p1",
                                           cond.target.short = "MUT",
                                           cond.ref.short    = "WT")
# p2
p2.mut.wt.res = analyse_diff_accessibility(cond.target.long  = "MUT_D11_p2",
                                           cond.ref.long     = "WT_D11_p2",
                                           cond.target.short = "MUT",
                                           cond.ref.short    = "WT")
# pMN
pM.mut.wt.res = analyse_diff_accessibility(cond.target.long  = "MUT_D11_pM",
                                           cond.ref.long     = "WT_D11_pM",
                                           cond.target.short = "MUT",
                                           cond.ref.short    = "WT")
```

Almost all significantly regulated regions are downregulated, which is exactly as expected with NFIA's function of opening chromatin in WT.

## Find regions with NFIA-dependent accessibility in WT-D11 vs WT-D7

For each domain, intersect the set of regions upregulated in WT-D11 vs WT-D7 with the set of regions downregulated in MUT-D11 vs WT-D11:

```{r, include=T}
# Regions upregulated in WT-D11 vs WT-D7
p1.11.7.res.up = as.data.frame(p1.11.7.res) %>% 
  filter(padj < fdr & abs(log2FoldChange) > min.l2fc & baseMean > min.baseMean) %>%
  filter(log2FoldChange > 0) %>% 
  rownames()

p2.11.7.res.up = as.data.frame(p2.11.7.res) %>% 
  filter(padj < fdr & abs(log2FoldChange) > min.l2fc & baseMean > min.baseMean) %>%
  filter(log2FoldChange > 0) %>% 
  rownames()

pM.11.7.res.up = as.data.frame(pM.11.7.res) %>% 
  filter(padj < fdr & abs(log2FoldChange) > min.l2fc & baseMean > min.baseMean) %>%
  filter(log2FoldChange > 0) %>% 
  rownames()

# Regions downregulated in MUT-D11 vs WT-D11
p1.mut.wt.res.down = as.data.frame(p1.mut.wt.res) %>% 
  filter(padj < fdr & abs(log2FoldChange) > min.l2fc & baseMean > min.baseMean) %>%
  filter(log2FoldChange < 0) %>% 
  rownames()

p2.mut.wt.res.down = as.data.frame(p2.mut.wt.res) %>% 
  filter(padj < fdr & abs(log2FoldChange) > min.l2fc & baseMean > min.baseMean) %>%
  filter(log2FoldChange < 0) %>% 
  rownames()

pM.mut.wt.res.down = as.data.frame(pM.mut.wt.res) %>% 
  filter(padj < fdr & abs(log2FoldChange) > min.l2fc & baseMean > min.baseMean) %>%
  filter(log2FoldChange < 0) %>% 
  rownames()

# NFIA-dependent regions upregulated in WT-D11 vs WT-D7
p1.11.7.dep = intersect(p1.11.7.res.up, p1.mut.wt.res.down)

p2.11.7.dep = intersect(p2.11.7.res.up, p2.mut.wt.res.down)

pM.11.7.dep = intersect(pM.11.7.res.up, pM.mut.wt.res.down)

# Save these sets of regions
saveRDS(p1.11.7.dep, file = paste0("../r_results/diff_accessibility/tables/p1_D11_vs_D7_dep", 
                                   "_fdr", fdr, "_min-l2fc", min.l2fc, "_min-baseMean", min.baseMean, ".rds"))

saveRDS(p2.11.7.dep, file = paste0("../r_results/diff_accessibility/tables/p2_D11_vs_D7_dep", 
                                   "_fdr", fdr, "_min-l2fc", min.l2fc, "_min-baseMean", min.baseMean, ".rds"))

saveRDS(pM.11.7.dep, file = paste0("../r_results/diff_accessibility/tables/pM_D11_vs_D7_dep", 
                                   "_fdr", fdr, "_min-l2fc", min.l2fc, "_min-baseMean", min.baseMean, ".rds"))
```

Print out the numbers of NFIA-dependent regions per domain:

```{r, include=T}
cat("Number of NFIA-dependent regions in p1:", length(p1.11.7.dep), "\n")

cat("Number of NFIA-dependent regions in p2:", length(p2.11.7.dep), "\n")

cat("Number of NFIA-dependent regions in pM:", length(pM.11.7.dep), "\n")
```
