---
title: "diff_expression"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
dyn.load("/home/rstudio/libs/libxml2.so.2")
library(DESeq2)
library(dplyr)
library(tibble)
library(stringr)
library(apeglm)
library(ggplot2)
dyn.load("/home/rstudio/libs/libgsl.so.25")
dyn.load("/home/rstudio/libs/libgslcblas.so.0")
library(hypeR)
```

## Init

Set FDR:

```{r, include=T}
fdr = 0.05
```

Define a general function for the differential expression analysis:

```{r, include=T}
diff_expression = function(p, p.cond, diff.levels, de.results.rds.filename) {
  p = round(p)
  
  row.names(p.cond) = names(p)
  
  p.cond$condition = factor(p.cond$condition, levels = diff.levels)
  
  p.dds = DESeqDataSetFromMatrix(countData = p,
                                 colData = p.cond,
                                 design = ~ condition)
  
  p.dds = DESeq(p.dds)
  
  p.res = results(p.dds, 
                  contrast = c("condition", diff.levels[1], diff.levels[2]), 
                  alpha = fdr)
  
  saveRDS(p.res, file = paste0("../r_results/diff_expression/tables/", de.results.rds.filename))
  
  de.results = list(p.res, p.dds)
  
  names(de.results) = c("results_object", "deseq_object")
  
  return(de.results)
}
```

Define a general function to gather differential expression stats:

```{r, include=T}
print_de_summary = function(p.res) {
  cat("Number of significantly differentially regulated genes (FDR =", fdr * 100, "%):", 
    sum(p.res$padj < fdr, na.rm = T), 
    "\n")

  p.n.up = nrow(as.data.frame(p.res) %>% 
    filter(!is.na(padj)) %>%
    filter(padj < fdr) %>%
    filter(log2FoldChange > 0))
  
  cat("Number of upregulated genes (FDR =", fdr * 100, "%):", 
      p.n.up,
      "\n")
  
  p.n.down = nrow(as.data.frame(p.res) %>% 
    filter(!is.na(padj)) %>%
    filter(padj < fdr) %>%
    filter(log2FoldChange < 0))
  
  cat("Number of downregulated genes (FDR =", fdr * 100, "%):", 
      p.n.down,
      "\n")
  
  p.n.all = nrow(as.data.frame(p.res))
  
  cat("The total number of genes:",
      p.n.all,
      "\n")
  
  cat("The percentage of significantly differentially regulated genes:",
      round((p.n.up + p.n.down) / p.n.all * 100, 2))
  
  deg.proportions = list(p.n.up, p.n.down, p.n.all)
  
  names(deg.proportions) = c("up", "down", "all")
  
  return(deg.proportions)
}
```

Define a general function to do PCA:

```{r, include=T}
deg_pca = function(p.dds, pca.plot.filename) {
  p.rld = rlog(p.dds)
  
  # Based on the code from DESeq2 vignette: https://bioconductor.org/packages/devel/bioc/vignettes/DESeq2/inst/doc/DESeq2.html#principal-component-plot-of-the-samples
  p.pcaData = plotPCA(p.rld, returnData = T)
  
  p.percentVar = round(100 * attr(p.pcaData, "percentVar"), 2)
  
  pca.plot = ggplot(p.pcaData, aes(PC1, PC2, color = condition)) +
    geom_point(size = 2) +
    xlab(paste0("PC1: ", p.percentVar[1], "% variance")) +
    ylab(paste0("PC2: ", p.percentVar[2], "% variance")) + 
    coord_fixed() +
    theme_classic()
  
  ggsave(filename = paste0("../r_results/diff_expression/plots/", pca.plot.filename),
         plot = pca.plot)
  
  return(pca.plot)
}
```

Define a general function for plotting the proportions of differentially expressed genes:

```{r, include=T}
plot_deg_proportions = function(p.deseq.object, deg.proportions.plot.filename) {
  p.n.up = p.deseq.object[["up"]]

  p.n.down = p.deseq.object[["down"]]
  
  p.n.all = p.deseq.object[["all"]]

  p.freq.up = p.n.up / p.n.all
  
  p.freq.down = p.n.down / p.n.all
  
  p.freq.nonsign = 1 - p.freq.up - p.freq.down
  
  p.freq.df = data.frame(freq         = c(p.freq.up, p.freq.down, p.freq.nonsign),
                         category     = c("Up",      "Down",      "Non-significant"),
                         element_type = "Genes") %>%
    dplyr::mutate(category = factor(category, levels = category))
  
  prop.plot = ggplot(p.freq.df) +
    geom_col(aes(x = element_type,
                 y = freq,
                 fill = category)) + 
    theme_classic()
  
  ggsave(filename = paste0("../r_results/diff_expression/plots/", deg.proportions.plot.filename),
         plot = prop.plot)
  
  return(prop.plot)
}
```

Load the master table:

```{r, include=T}
master.table = assays(readRDS("../input/salmon.merged.gene_counts.rds"))$counts
```

## Generate the reference set of genes in p1 (direct and indirect targets)

1) Find differentially expressed genes in p1 in D11-MUT vs D11-WT:

```{r, include=T}
p1.mut.wt.raw = master.table %>%
  dplyr::select(MUT_D11_p1_NFIAn_R2,
                MUT_D11_p1_NFIAn_R3,
                WT_D11_p1_NFIAp_R2,
                WT_D11_p1_NFIAp_R3)

# A command like 
# unlist(stringr::str_match_all(names(master.table), "MUT_D11_p1.*")) 
# is convenient for extracting sample names for a given condition 
# from the header of the master table.

p1.mut.wt.cond.raw = data.frame(condition = stringr::str_extract(names(p1.mut.wt.raw), "(MUT|WT)"))

diff.levels.mut.wt = c("MUT", "WT")

p1.mut.wt.results = diff_expression(p1.mut.wt.raw, p1.mut.wt.cond.raw, diff.levels.mut.wt, "deseq-results_p1-MUT-D11-NFIAn_vs_p1-WT-D11-NFIAp.rds")

p1.mut.wt.res = p1.mut.wt.results[["results_object"]]
```

Print a summary of the results:

```{r, include=T}
p1.deg_proportions = print_de_summary(p1.mut.wt.res)
```

Do PCA:

```{r, include=T}
p1.mut.wt.pca = deg_pca(p1.mut.wt.results[["deseq_object"]], "pca_diff-expression_p1-MUT-D11-NFIAn_vs_p1-WT-D11-NFIAp.pdf")

# p1.mut.wt.pca
```

The main difference is not between the conditions, but between one WT sample and the other three samples, which explains the small number of differentially expressed genes.

2) Find differentially expressed genes in p1 in D11-WT vs D7-WT:

```{r, include=T}
p1.11.7.raw = master.table %>%
  dplyr::select(WT_D11_p1_NFIAp_R2,
                WT_D11_p1_NFIAp_R3,
                WT_D7_p1_NFIAn_R1,
                WT_D7_p1_NFIAn_R2,
                WT_D7_p1_NFIAn_R3)

p1.11.7.cond.raw = data.frame(condition = stringr::str_extract(names(p1.11.7.raw), "(D11|D7)"))

diff.levels = c("D11", "D7")

p1.11.7.results = diff_expression(p1.11.7.raw, p1.11.7.cond.raw, diff.levels, "deseq-results_p1-WT-D11-NFIAp_vs_p1-WT-D7-NFIAn.rds")

p1.11.7.res = p1.11.7.results[["results_object"]]
```

Print a summary of the results:

```{r, include=T}
p1.11.7.deg_proportions = print_de_summary(p1.11.7.res)
```

Do PCA:

```{r, include=T}
p1.11.7.pca = deg_pca(p1.11.7.results[["deseq_object"]], "pca_diff-expression_p1-WT-D11-NFIAp_vs_p1-WT-D7-NFIAn.pdf")

# p1.11.7.pca
```

The two conditions are very well separated by PC1.

Check the expression of Nfia across the two conditions:

```{r, include=T}
p1.11.7.dds = p1.11.7.results[["deseq_object"]]

p1.11.7.dds = estimateSizeFactors(p1.11.7.dds)

p1.11.7.norm.counts = counts(p1.11.7.dds, normalized = T)

p1.11.7.norm.counts %>%
  as.data.frame() %>%
  rownames_to_column(var = "gene_name") %>%
  filter(gene_name == "Nfia") %>%
  left_join(as.data.frame(p1.11.7.res) %>%
              rownames_to_column(var = "gene_name"),
            by = c("gene_name" = "gene_name")) %>%
  dplyr::select(-baseMean,
                -lfcSE,
                -stat,
                -pvalue)
```

Exactly as expected, Nfia is significantly upregulated.

3) Intersect the set of DEGs from MUT-D11 vs WT-D11 with the set of DEGs from WT-D11 vs WT-D7 to obtain the p1-specific set of reference genes. A reference gene should be:

a) either significantly downregulated in MUT-D11 vs WT-D11 but significantly upregulated in WT-D11 vs WT-D7,

b) or significantly upregulated in MUT-D11 vs WT-D11 but significantly downregulated in WT-D11 vs WT-D7:

```{r, include=T}
p1.degs.mut.wt.up = as.data.frame(p1.mut.wt.res) %>%
  rownames_to_column(var = "gene_name") %>%
  filter(padj < fdr) %>%
  filter(log2FoldChange > 0) %>%
  pull(gene_name)

p1.degs.mut.wt.down = as.data.frame(p1.mut.wt.res) %>%
  rownames_to_column(var = "gene_name") %>%
  filter(padj < fdr) %>%
  filter(log2FoldChange < 0) %>%
  pull(gene_name)

p1.degs.11.7.up = as.data.frame(p1.11.7.res) %>%
  rownames_to_column(var = "gene_name") %>%
  filter(padj < fdr) %>%
  filter(log2FoldChange > 0) %>%
  pull(gene_name)

p1.degs.11.7.down = as.data.frame(p1.11.7.res) %>%
  rownames_to_column(var = "gene_name") %>%
  filter(padj < fdr) %>%
  filter(log2FoldChange < 0) %>%
  pull(gene_name)

p1.ref.genes = union(intersect(p1.degs.mut.wt.down, p1.degs.11.7.up),
                     intersect(p1.degs.mut.wt.up, p1.degs.11.7.down))

cat("Number of p1 reference genes:", length(p1.ref.genes), "\n")

saveRDS(p1.ref.genes, file = "../r_results/diff_expression/tables/p1_ref_genes.rds")
```

## Generate the reference set of genes in p2 (direct and indirect targets)

1) Find differentially expressed genes in p2 in D11-MUT vs D11-WT:

```{r, include=T}
p2.mut.wt.raw = master.table %>%
  dplyr::select(MUT_D11_p2_NFIAn_R2,
                MUT_D11_p2_NFIAn_R3,
                WT_D11_p2_NFIAp_R2,
                WT_D11_p2_NFIAp_R3)

p2.mut.wt.cond.raw = data.frame(condition = stringr::str_extract(names(p2.mut.wt.raw), "(MUT|WT)"))

diff.levels = c("MUT", "WT")

p2.mut.wt.results = diff_expression(p2.mut.wt.raw, p2.mut.wt.cond.raw, diff.levels, "deseq-results_p2-MUT-D11-NFIAn_vs_p2-WT-D11-NFIAp.rds")

p2.mut.wt.res = p2.mut.wt.results[["results_object"]]
```

Print a summary of the results:

```{r, include=T}
p2.mut.wt.deg_proportions = print_de_summary(p2.mut.wt.res)
```

Do PCA:

```{r, include=T}
p2.mut.wt.pca = deg_pca(p2.mut.wt.results[["deseq_object"]], "pca_diff-expression_p2-MUT-D11-NFIAn_vs_p2-WT-D11-NFIAp.pdf")

# p2.mut.wt.pca
```

The conditions are separated by PC2, but it explains only 22% of the variance, and the samples from the same condition are still quite far apart; hence, the minute number of differentially expressed genes.

2) Find differentially expressed genes in p2 in D11-WT vs D7-WT:

```{r, include=T}
p2.11.7.raw = master.table %>%
  dplyr::select(WT_D11_p2_NFIAp_R2,
                WT_D11_p2_NFIAp_R3,
                WT_D7_p2_NFIAn_R1,
                WT_D7_p2_NFIAn_R2,
                WT_D7_p2_NFIAn_R3)

p2.11.7.cond.raw = data.frame(condition = stringr::str_extract(names(p2.11.7.raw), "(D11|D7)"))

diff.levels = c("D11", "D7")

p2.11.7.results = diff_expression(p2.11.7.raw, p2.11.7.cond.raw, diff.levels, "deseq-results_p2-WT-D11-NFIAp_vs_p2-WT-D7-NFIAn.rds")

p2.11.7.res = p2.11.7.results[["results_object"]]
```

Print a summary of the results:

```{r, include=T}
p2.11.7.deg_proportions = print_de_summary(p2.11.7.res)
```

Do PCA:

```{r, include=T}
p2.11.7.pca = deg_pca(p2.11.7.results[["deseq_object"]], "pca_diff-expression_p2-WT-D11-NFIAp_vs_p2-WT-D7-NFIAn.pdf")

# p2.11.7.pca
```

The two conditions are very well separated by PC1.

Check the expression of Nfia across the two conditions:

```{r, include=T}
p2.11.7.dds = p2.11.7.results[["deseq_object"]]

p2.11.7.dds = estimateSizeFactors(p2.11.7.dds)

p2.11.7.norm.counts = counts(p2.11.7.dds, normalized = T)

p2.11.7.norm.counts %>%
  as.data.frame() %>%
  rownames_to_column(var = "gene_name") %>%
  filter(gene_name == "Nfia") %>%
  left_join(as.data.frame(p2.11.7.res) %>%
              rownames_to_column(var = "gene_name"),
            by = c("gene_name" = "gene_name")) %>%
  dplyr::select(-baseMean,
                -lfcSE,
                -stat,
                -pvalue)
```

Exactly as expected, Nfia is significantly upregulated. 

3) Intersect the set of DEGs from MUT-D11 vs WT-D11 with the set of DEGs from WT-D11 vs WT-D7 to obtain the p2-specific set of reference genes. A reference gene should be:

a) either significantly downregulated in MUT-D11 vs WT-D11 but significantly upregulated in WT-D11 vs WT-D7,

b) or significantly upregulated in MUT-D11 vs WT-D11 but significantly downregulated in WT-D11 vs WT-D7:

```{r, include=T}
p2.degs.mut.wt.up = as.data.frame(p2.mut.wt.res) %>%
  rownames_to_column(var = "gene_name") %>%
  filter(padj < fdr) %>%
  filter(log2FoldChange > 0) %>%
  pull(gene_name)

p2.degs.mut.wt.down = as.data.frame(p2.mut.wt.res) %>%
  rownames_to_column(var = "gene_name") %>%
  filter(padj < fdr) %>%
  filter(log2FoldChange < 0) %>%
  pull(gene_name)

p2.degs.11.7.up = as.data.frame(p2.11.7.res) %>%
  rownames_to_column(var = "gene_name") %>%
  filter(padj < fdr) %>%
  filter(log2FoldChange > 0) %>%
  pull(gene_name)

p2.degs.11.7.down = as.data.frame(p2.11.7.res) %>%
  rownames_to_column(var = "gene_name") %>%
  filter(padj < fdr) %>%
  filter(log2FoldChange < 0) %>%
  pull(gene_name)

p2.ref.genes = union(intersect(p2.degs.mut.wt.down, p2.degs.11.7.up),
                     intersect(p2.degs.mut.wt.up, p2.degs.11.7.down))

cat("Number of p2 reference genes:", length(p2.ref.genes), "\n")

saveRDS(p2.ref.genes, file = "../r_results/diff_expression/tables/p2_ref_genes.rds")
```

## Generate the reference set of genes in pMN (direct and indirect targets)

1) Find differentially expressed genes in pMN in D11-MUT vs D11-WT:

```{r, include=T}
pM.mut.wt.raw = master.table %>%
  dplyr::select(MUT_D11_pM_NFIAn_R2,
                MUT_D11_pM_NFIAn_R3,
                WT_D11_pM_NFIAp_R2,
                WT_D11_pM_NFIAp_R3)

pM.mut.wt.cond.raw = data.frame(condition = stringr::str_extract(names(pM.mut.wt.raw), "(MUT|WT)"))

diff.levels = c("MUT", "WT")

pM.mut.wt.results = diff_expression(pM.mut.wt.raw, pM.mut.wt.cond.raw, diff.levels, "deseq-results_pM-MUT-D11-NFIAn_vs_pM-WT-D11-NFIAp.rds")

pM.mut.wt.res = pM.mut.wt.results[["results_object"]]
```

Print a summary of the results:

```{r, include=T}
pM.mut.wt.deg_proportions = print_de_summary(pM.mut.wt.res)
```

Do PCA:

```{r, include=T}
pM.mut.wt.pca = deg_pca(pM.mut.wt.results[["deseq_object"]], "pca_diff-expression_pM-MUT-D11-NFIAn_vs_pM-WT-D11-NFIAp.pdf")

# pM.pca
```

The conditions are separated by PC1, and it explains 71% of the variance; however, the samples from the same condition are still quite far apart from each other. Hence, although there are more differentially expressed genes in pMNs, than in p1 or p2, but still not many.

2) Find differentially expressed genes in pMN in D11-WT vs D7-WT:

```{r, include=T}
pM.11.7.raw = master.table %>%
  dplyr::select(WT_D11_pM_NFIAp_R2,
                WT_D11_pM_NFIAp_R3,
                WT_D7_pM_NFIAn_R1,
                WT_D7_pM_NFIAn_R2,
                WT_D7_pM_NFIAn_R3)

pM.11.7.cond.raw = data.frame(condition = stringr::str_extract(names(pM.11.7.raw), "(D11|D7)"))

diff.levels = c("D11", "D7")

pM.11.7.results = diff_expression(pM.11.7.raw, pM.11.7.cond.raw, diff.levels, "deseq-results_pM-WT-D11-NFIAp_vs_pM-WT-D7-NFIAn.rds")

pM.11.7.res = pM.11.7.results[["results_object"]]
```

Print a summary of the results:

```{r, include=T}
pM.11.7.deg_proportions = print_de_summary(pM.11.7.res)
```

Do PCA:

```{r, include=T}
pM.11.7.pca = deg_pca(pM.11.7.results[["deseq_object"]], "pca_diff-expression_pM-WT-D11-NFIAp_vs_pM-WT-D7-NFIAn.pdf")

# p1.11.7.pca
```

The two conditions are very well separated by PC1.

Check the expression of Nfia across the two conditions:

```{r, include=T}
pM.11.7.dds = pM.11.7.results[["deseq_object"]]

pM.11.7.dds = estimateSizeFactors(pM.11.7.dds)

pM.11.7.norm.counts = counts(pM.11.7.dds, normalized = T)

pM.11.7.norm.counts %>%
  as.data.frame() %>%
  rownames_to_column(var = "gene_name") %>%
  filter(gene_name == "Nfia") %>%
  left_join(as.data.frame(pM.11.7.res) %>%
              rownames_to_column(var = "gene_name"),
            by = c("gene_name" = "gene_name")) %>%
  dplyr::select(-baseMean,
                -lfcSE,
                -stat,
                -pvalue)
```

Exactly as expected, Nfia is significantly upregulated.

3) Intersect the set of DEGs from MUT-D11 vs WT-D11 with the set of DEGs from WT-D11 vs WT-D7 to obtain the pMN-specific set of reference genes: a reference gene should be significantly downregulated in MUT-D11 vs WT-D11 but significantly upregulated in WT-D11 vs WT-D7:

```{r, include=T}
pM.degs.mut.wt.up = as.data.frame(pM.mut.wt.res) %>%
  rownames_to_column(var = "gene_name") %>%
  filter(padj < fdr) %>%
  filter(log2FoldChange > 0) %>%
  pull(gene_name)

pM.degs.mut.wt.down = as.data.frame(pM.mut.wt.res) %>%
  rownames_to_column(var = "gene_name") %>%
  filter(padj < fdr) %>%
  filter(log2FoldChange < 0) %>%
  pull(gene_name)

pM.degs.11.7.up = as.data.frame(pM.11.7.res) %>%
  rownames_to_column(var = "gene_name") %>%
  filter(padj < fdr) %>%
  filter(log2FoldChange > 0) %>%
  pull(gene_name)

pM.degs.11.7.down = as.data.frame(pM.11.7.res) %>%
  rownames_to_column(var = "gene_name") %>%
  filter(padj < fdr) %>%
  filter(log2FoldChange < 0) %>%
  pull(gene_name)

pM.ref.genes = union(intersect(pM.degs.mut.wt.down, pM.degs.11.7.up),
                     intersect(pM.degs.mut.wt.up, pM.degs.11.7.down))

cat("Number of pM reference genes:", length(pM.ref.genes), "\n")

saveRDS(pM.ref.genes, file = "../r_results/diff_expression/tables/pM_ref_genes.rds")
```

## Check known NFIA targets (direct or indirect) among the reference genes

Check known NFIA targets in the developing spinal cord among the reference genes for each domains.

Load the names of the NFIA target genes from the spinal cord:

```{r, include=T}
spinal.cord.targets = read.delim("../input/NFIA_regulated_genes_in_spinal_cord.tsv",
                                 header = F)

names(spinal.cord.targets) = c("gene_name", "regulation_mode", "ref", "pmid")
```

a) Check the spinal cord targets among the p1-specific reference genes:

```{r, include=T}
p1.ref.genes[p1.ref.genes %in% spinal.cord.targets$gene_name]
```

b) Check the spinal cord targets among the p2-specific reference genes:

```{r, include=T}
p2.ref.genes[p2.ref.genes %in% spinal.cord.targets$gene_name]
```

c) Check the spinal cord targets among the pMN-specific reference genes:

```{r, include=T}
pM.ref.genes[pM.ref.genes %in% spinal.cord.targets$gene_name]
```
