---
title: "predict_correlated_deg"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
dyn.load("/home/rstudio/libs/libxml2.so.2")
library(DESeq2)
library(dplyr)
library(GenomicRanges)
library(tibble)
library(rtracklayer)
library(ggplot2)
library(stringr)
library(scales)
```

Define constants:

```{r, include=T}
pcc.cutoff = 0.9

fdr.matches = 0.05

vicinity.radius = 500 # kbp

min.tad.size = 500 # kbp

max.tad.size = 1500 # kbp
```

Define cutoff sets:

```{r, include=T}
region.cutoffs = list("set1" = list("fdr" = 0.01, "min.l2fc" = 2, "min.baseMean" = 100),
                      "set2" = list("fdr" = 0.01, "min.l2fc" = 2, "min.baseMean" = 80),
                      "set3" = list("fdr" = 0.01, "min.l2fc" = 2, "min.baseMean" = 60),
                      "set4" = list("fdr" = 0.01, "min.l2fc" = 2, "min.baseMean" = 40),
                      "set5" = list("fdr" = 0.01, "min.l2fc" = 2, "min.baseMean" = 20),
                      "set6" = list("fdr" = 0.01, "min.l2fc" = 1.5, "min.baseMean" = 20),
                      "set7" = list("fdr" = 0.01, "min.l2fc" = 1, "min.baseMean" = 20),
                      "set8" = list("fdr" = 0.01, "min.l2fc" = 0.5, "min.baseMean" = 20),
                      "set9" = list("fdr" = 0.02, "min.l2fc" = 0.5, "min.baseMean" = 20),
                      "set10" = list("fdr" = 0.03, "min.l2fc" = 0.5, "min.baseMean" = 20),
                      "set11" = list("fdr" = 0.04, "min.l2fc" = 0.5, "min.baseMean" = 20),
                      "set12" = list("fdr" = 0.05, "min.l2fc" = 0.5, "min.baseMean" = 20))
```

Define a general function to select predicted NFIA-dependent region-gene pairs:

```{r, include=T}
select_predicted_pairs = function(p.corr, base.filename) {
  p.corr.predicted = p.corr %>%
    filter(abs_pcc > pcc.cutoff)
  
  saveRDS(p.corr.predicted,
          paste0("../r_results/predict_correlated_deg_highest/tables/",
                 base.filename, ".rds"))
  
  write.table(p.corr.predicted,
              file = paste0("../r_results/predict_correlated_deg_highest/tables/",
                            base.filename, ".tsv"),
              quote = F,
              sep = "\t",
              row.names = F,
              col.names = F)
  
  return(p.corr.predicted)
}
```

Define a general function to find the proportions of shared and unique features in three sets:

```{r, include=T}
find_shared_features = function(p1, p2, pM, plot.filename) {
  union.all = unique(c(p1, p2, pM))
  
  union.12 = unique(c(p1, p2))
  
  union.1M = unique(c(p1, pM))
  
  union.2M = unique(c(p2, pM))
  
  inter.all = intersect(intersect(p1, p2), pM)
  
  inter.12 = intersect(p1, p2)
  
  inter.1M = intersect(p1, pM)
  
  inter.2M = intersect(p2, pM)
  
  proportions.df = data.frame(p1_num = length(p1),
                              p2_num = length(p2),
                              pM_num = length(pM),
                              p1 = length(setdiff(union.all, union.2M)) / length(p1),
                              p2 = length(setdiff(union.all, union.1M)) / length(p2),
                              pM = length(setdiff(union.all, union.12)) / length(pM),
                              p12_1 = length(setdiff(inter.12, inter.all)) / length(p1),
                              p12_2 = length(setdiff(inter.12, inter.all)) / length(p2),
                              p1M_1 = length(setdiff(inter.1M, inter.all)) / length(p1),
                              p1M_M = length(setdiff(inter.1M, inter.all)) / length(pM),
                              p2M_2 = length(setdiff(inter.2M, inter.all)) / length(p2),
                              p2M_M = length(setdiff(inter.2M, inter.all)) / length(pM),
                              p12M_1 = length(inter.all) / length(p1),
                              p12M_2 = length(inter.all) / length(p2),
                              p12M_M = length(inter.all) / length(pM))
  
  p = data.frame(category = c("1", "2", "M",
                              rep("12", 2), rep("1M", 2), rep("2M", 2),
                              rep("12M", 3)),
                 domain = c("p1", "p2", "pM",
                            "p1", "p2", "p1", "pM", "p2", "pM",
                            "p1", "p2", "pM"),
                 proportion = c(proportions.df$p1,
                                proportions.df$p2,
                                proportions.df$pM,
                                proportions.df$p12_1,
                                proportions.df$p12_2,
                                proportions.df$p1M_1,
                                proportions.df$p1M_M,
                                proportions.df$p2M_2,
                                proportions.df$p2M_M,
                                proportions.df$p12M_1,
                                proportions.df$p12M_2,
                                proportions.df$p12M_M)) %>% 
    mutate(category = factor(category, levels = c("1", "2", "M",
                                                  "12", "1M", "2M",
                                                  "12M"))) %>%
    ggplot(aes(x = category, y = proportion, fill = domain)) +
      geom_col(position = position_dodge2(preserve = "single")) +
      theme_classic()
  
  ggsave(filename = plot.filename,
         plot = p)  
  
  return(proportions.df)
}
```

Find how many NFIA-dependent region-gene pairs calculated in radius-based vicinities have absolute PCCs (hereafter called just PCCs for simplicity) above the cutoff:

```{r, include=T}
num.top.pairs.radius.df = bind_rows(lapply(names(region.cutoffs), 
                                           function(s) {
                                             fdr = region.cutoffs[[s]]$fdr
                                             min.l2fc = region.cutoffs[[s]]$min.l2fc
                                             min.baseMean = region.cutoffs[[s]]$min.baseMean
                                             p1.corr = readRDS(paste0("../r_results/predict_correlated_deg/p1_corr_in_radius_", 
                                                                      vicinity.radius, "kbp_dep_regions_fdr", fdr, "_min-l2fc", min.l2fc,
                                                                      "_min-baseMean", min.baseMean, ".rds"))
                                             p2.corr = readRDS(paste0("../r_results/predict_correlated_deg/p2_corr_in_radius_", 
                                                                      vicinity.radius, "kbp_dep_regions_fdr", fdr, "_min-l2fc", min.l2fc,
                                                                      "_min-baseMean", min.baseMean, ".rds"))
                                             pM.corr = readRDS(paste0("../r_results/predict_correlated_deg/pM_corr_in_radius_", 
                                                                      vicinity.radius, "kbp_dep_regions_fdr", fdr, "_min-l2fc", min.l2fc,
                                                                      "_min-baseMean", min.baseMean, ".rds"))
                                             return(data.frame(cutoff_set = s,
                                                               p1 = nrow(p1.corr %>% filter(abs_pcc > pcc.cutoff)),
                                                               p2 = nrow(p2.corr %>% filter(abs_pcc > pcc.cutoff)),
                                                               pM = nrow(pM.corr %>% filter(abs_pcc > pcc.cutoff))))
                                           }))

p = data.frame(cutoff_set = rep(num.top.pairs.radius.df$cutoff_set, 3),
               domain = c(rep("p1", length(num.top.pairs.radius.df$cutoff_set)),
                          rep("p2", length(num.top.pairs.radius.df$cutoff_set)),
                          rep("pM", length(num.top.pairs.radius.df$cutoff_set))),
               num = c(num.top.pairs.radius.df$p1,
                       num.top.pairs.radius.df$p2,
                       num.top.pairs.radius.df$pM)) %>%
  mutate(cutoff_set = factor(cutoff_set, levels = num.top.pairs.radius.df$cutoff_set),
         domain = factor(domain, levels = c("p1", "p2", "pM"))) %>%
  ggplot(aes(x = cutoff_set,
             y = num,
             fill = domain)) +
    geom_col(position = "dodge") +
    scale_y_continuous(breaks = pretty_breaks(n = 14)) + 
    theme_classic()

ggsave(filename = "../r_results/predict_correlated_deg_highest/plots/numbers_of_region-gene_pairs_radius.pdf",
       plot = p) 
```

Find how many NFIA-dependent region-gene pairs calculated in size-selected TADs have PCCs above the cutoff:

```{r, include=T}
num.top.pairs.tad.df = bind_rows(lapply(names(region.cutoffs), 
                                        function(s) {
                                          fdr = region.cutoffs[[s]]$fdr
                                          min.l2fc = region.cutoffs[[s]]$min.l2fc
                                          min.baseMean = region.cutoffs[[s]]$min.baseMean
                                          p1.corr = readRDS(paste0("../r_results/predict_correlated_deg/p1_corr_in_largest_tad_",
                                                                   min.tad.size, "-", max.tad.size, "kbp_dep_regions_fdr", fdr, 
                                                                   "_min-l2fc", min.l2fc, "_min-baseMean", min.baseMean, ".rds"))
                                          p2.corr = readRDS(paste0("../r_results/predict_correlated_deg/p2_corr_in_largest_tad_",
                                                                   min.tad.size, "-", max.tad.size, "kbp_dep_regions_fdr", fdr, 
                                                                   "_min-l2fc", min.l2fc, "_min-baseMean", min.baseMean, ".rds"))
                                          pM.corr = readRDS(paste0("../r_results/predict_correlated_deg/pM_corr_in_largest_tad_",
                                                                   min.tad.size, "-", max.tad.size, "kbp_dep_regions_fdr", fdr, 
                                                                   "_min-l2fc", min.l2fc, "_min-baseMean", min.baseMean, ".rds"))
                                          return(data.frame(cutoff_set = s,
                                                            p1 = nrow(p1.corr %>% filter(abs_pcc > pcc.cutoff)),
                                                            p2 = nrow(p2.corr %>% filter(abs_pcc > pcc.cutoff)),
                                                            pM = nrow(pM.corr %>% filter(abs_pcc > pcc.cutoff))))
                                        }))

p = data.frame(cutoff_set = rep(num.top.pairs.tad.df$cutoff_set, 3),
               domain = c(rep("p1", length(num.top.pairs.tad.df$cutoff_set)),
                          rep("p2", length(num.top.pairs.tad.df$cutoff_set)),
                          rep("pM", length(num.top.pairs.tad.df$cutoff_set))),
               num = c(num.top.pairs.tad.df$p1,
                       num.top.pairs.tad.df$p2,
                       num.top.pairs.tad.df$pM)) %>%
  mutate(cutoff_set = factor(cutoff_set, levels = num.top.pairs.tad.df$cutoff_set),
         domain = factor(domain, levels = c("p1", "p2", "pM"))) %>%
  ggplot(aes(x = cutoff_set,
             y = num,
             fill = domain)) +
    geom_col(position = "dodge") +
    scale_y_continuous(breaks = pretty_breaks(n = 6)) + 
    theme_classic()

ggsave(filename = "../r_results/predict_correlated_deg_highest/plots/numbers_of_region-gene_pairs_tad.pdf",
       plot = p) 
```

In both cases, pMN dominates in terms of the number of predicted NFIA-dependent region-gene pairs. Additionally, the fastest growth of this number occurs when I decrease the minimum log2FoldChange. Hence, for the radius-based predictions, I selected two sets of cutoffs: a "strict" one (FDR = 1%, min{log2(FC)} = 2, min{baseMean} = 40) and a "permissive" one (FDR = 1%, min{log2(FC)} = 0.5, min{baseMean} = 20) -- roughly, at the boundaries of the rapid growth zone. For the TAD-based predictions, I selected only the permissive set, as otherwise I would have too few pairs from p1.

Form the sets of predicted NFIA-dependent region-gene pairs according to the selected sets of cutoffs:

```{r, include=T}
# Radius-based predictions

# Strict set: set 4

fdr = region.cutoffs[["set4"]]$fdr

min.l2fc = region.cutoffs[["set4"]]$min.l2fc

min.baseMean = region.cutoffs[["set4"]]$min.baseMean

p1.corr = readRDS(paste0("../r_results/predict_correlated_deg/p1_corr_in_radius_", 
                         vicinity.radius, "kbp_dep_regions_fdr", fdr, "_min-l2fc", min.l2fc,
                         "_min-baseMean", min.baseMean, ".rds"))

p2.corr = readRDS(paste0("../r_results/predict_correlated_deg/p2_corr_in_radius_", 
                         vicinity.radius, "kbp_dep_regions_fdr", fdr, "_min-l2fc", min.l2fc,
                         "_min-baseMean", min.baseMean, ".rds"))

pM.corr = readRDS(paste0("../r_results/predict_correlated_deg/pM_corr_in_radius_", 
                         vicinity.radius, "kbp_dep_regions_fdr", fdr, "_min-l2fc", min.l2fc,
                         "_min-baseMean", min.baseMean, ".rds"))
                                          
p1.predictions.radius.strict = select_predicted_pairs(p1.corr,
                                                      paste0("p1_predictions_corr_in_radius_", 
                                                             vicinity.radius, "kbp_dep_regions_fdr", fdr, "_min-l2fc", min.l2fc,
                                                             "_min-baseMean", min.baseMean))

p2.predictions.radius.strict = select_predicted_pairs(p2.corr,
                                                      paste0("p2_predictions_corr_in_radius_", 
                                                             vicinity.radius, "kbp_dep_regions_fdr", fdr, "_min-l2fc", min.l2fc,
                                                             "_min-baseMean", min.baseMean))

pM.predictions.radius.strict = select_predicted_pairs(pM.corr,
                                                      paste0("pM_predictions_corr_in_radius_", 
                                                             vicinity.radius, "kbp_dep_regions_fdr", fdr, "_min-l2fc", min.l2fc,
                                                             "_min-baseMean", min.baseMean))

# Permissive set: set 8

fdr = region.cutoffs[["set8"]]$fdr

min.l2fc = region.cutoffs[["set8"]]$min.l2fc

min.baseMean = region.cutoffs[["set8"]]$min.baseMean

p1.corr = readRDS(paste0("../r_results/predict_correlated_deg/p1_corr_in_radius_", 
                         vicinity.radius, "kbp_dep_regions_fdr", fdr, "_min-l2fc", min.l2fc,
                         "_min-baseMean", min.baseMean, ".rds"))

p2.corr = readRDS(paste0("../r_results/predict_correlated_deg/p2_corr_in_radius_", 
                         vicinity.radius, "kbp_dep_regions_fdr", fdr, "_min-l2fc", min.l2fc,
                         "_min-baseMean", min.baseMean, ".rds"))

pM.corr = readRDS(paste0("../r_results/predict_correlated_deg/pM_corr_in_radius_", 
                         vicinity.radius, "kbp_dep_regions_fdr", fdr, "_min-l2fc", min.l2fc,
                         "_min-baseMean", min.baseMean, ".rds"))
                                          
p1.predictions.radius.permissive = select_predicted_pairs(p1.corr,
                                                          paste0("p1_predictions_corr_in_radius_", 
                                                                 vicinity.radius, "kbp_dep_regions_fdr", fdr, "_min-l2fc", min.l2fc,
                                                                 "_min-baseMean", min.baseMean))

p2.predictions.radius.permissive = select_predicted_pairs(p2.corr,
                                                          paste0("p2_predictions_corr_in_radius_", 
                                                                 vicinity.radius, "kbp_dep_regions_fdr", fdr, "_min-l2fc", min.l2fc,
                                                                 "_min-baseMean", min.baseMean))

pM.predictions.radius.permissive = select_predicted_pairs(pM.corr,
                                                          paste0("pM_predictions_corr_in_radius_", 
                                                                 vicinity.radius, "kbp_dep_regions_fdr", fdr, "_min-l2fc", min.l2fc,
                                                                 "_min-baseMean", min.baseMean))

# TAD-based predictions

# Permissive set: set 8

fdr = region.cutoffs[["set8"]]$fdr

min.l2fc = region.cutoffs[["set8"]]$min.l2fc

min.baseMean = region.cutoffs[["set8"]]$min.baseMean

p1.corr = readRDS(paste0("../r_results/predict_correlated_deg/p1_corr_in_largest_tad_",
                         min.tad.size, "-", max.tad.size, "kbp_dep_regions_fdr", fdr,
                         "_min-l2fc", min.l2fc, "_min-baseMean", min.baseMean, ".rds"))

p2.corr = readRDS(paste0("../r_results/predict_correlated_deg/p2_corr_in_largest_tad_",
                         min.tad.size, "-", max.tad.size, "kbp_dep_regions_fdr", fdr,
                         "_min-l2fc", min.l2fc, "_min-baseMean", min.baseMean, ".rds"))

pM.corr = readRDS(paste0("../r_results/predict_correlated_deg/pM_corr_in_largest_tad_",
                         min.tad.size, "-", max.tad.size, "kbp_dep_regions_fdr", fdr,
                         "_min-l2fc", min.l2fc, "_min-baseMean", min.baseMean, ".rds"))
                                          
p1.predictions.tad.permissive = select_predicted_pairs(p1.corr,
                                                       paste0("p1_predictions_corr_in_largest_tad_", 
                                                              min.tad.size, "-", max.tad.size, "kbp_dep_regions_fdr", 
                                                              fdr, "_min-l2fc", min.l2fc,
                                                              "_min-baseMean", min.baseMean))

p2.predictions.tad.permissive = select_predicted_pairs(p2.corr,
                                                       paste0("p2_predictions_corr_in_largest_tad_", 
                                                              min.tad.size, "-", max.tad.size, "kbp_dep_regions_fdr", 
                                                              fdr, "_min-l2fc", min.l2fc,
                                                              "_min-baseMean", min.baseMean))

pM.predictions.tad.permissive = select_predicted_pairs(pM.corr,
                                                       paste0("pM_predictions_corr_in_largest_tad_", 
                                                              min.tad.size, "-", max.tad.size, "kbp_dep_regions_fdr", 
                                                              fdr, "_min-l2fc", min.l2fc,
                                                              "_min-baseMean", min.baseMean))
```

Stratify the obtained predicted NFIA-dependent region-gene pairs into those with an NFIA element (putative direct regulation) and others (putative indirect regulation):

```{r, include=T, fig.width=3}
# Radius-based predictions

# Strict set: set 4

fdr = region.cutoffs[["set4"]]$fdr

min.l2fc = region.cutoffs[["set4"]]$min.l2fc

min.baseMean = region.cutoffs[["set4"]]$min.baseMean

p1.elements = names(readRDS(paste0("../r_results/select_diff_regions/", 
                                   "p1_dep_nfia_ranges_with_sites_plus-strand_80pc_match-fdr_", fdr.matches, 
                                   "_in-regions-of_fdr", fdr, "_min-l2fc", min.l2fc, "_min-baseMean", min.baseMean, ".rds")))

p2.elements = names(readRDS(paste0("../r_results/select_diff_regions/", 
                                   "p2_dep_nfia_ranges_with_sites_plus-strand_80pc_match-fdr_", fdr.matches, 
                                   "_in-regions-of_fdr", fdr, "_min-l2fc", min.l2fc, "_min-baseMean", min.baseMean, ".rds")))

pM.elements = names(readRDS(paste0("../r_results/select_diff_regions/", 
                                   "pM_dep_nfia_ranges_with_sites_plus-strand_80pc_match-fdr_", fdr.matches, 
                                   "_in-regions-of_fdr", fdr, "_min-l2fc", min.l2fc, "_min-baseMean", min.baseMean, ".rds")))
    
p1.predictions.radius.strict.strat = p1.predictions.radius.strict %>% 
  rowwise() %>%
  mutate(is_element = (region_id %in% p1.elements)) %>%
  ungroup()

p2.predictions.radius.strict.strat = p2.predictions.radius.strict %>% 
  rowwise() %>%
  mutate(is_element = (region_id %in% p2.elements)) %>%
  ungroup()

pM.predictions.radius.strict.strat = pM.predictions.radius.strict %>% 
  rowwise() %>%
  mutate(is_element = (region_id %in% pM.elements)) %>%
  ungroup()

predictions.radius.strict.strat = c(sum(p1.predictions.radius.strict.strat$is_element) / nrow(p1.predictions.radius.strict.strat),
                                    1 - sum(p1.predictions.radius.strict.strat$is_element) / nrow(p1.predictions.radius.strict.strat),
                                    sum(p2.predictions.radius.strict.strat$is_element) / nrow(p2.predictions.radius.strict.strat),
                                    1 - sum(p2.predictions.radius.strict.strat$is_element) / nrow(p2.predictions.radius.strict.strat),
                                    sum(pM.predictions.radius.strict.strat$is_element) / nrow(pM.predictions.radius.strict.strat),
                                    1 - sum(pM.predictions.radius.strict.strat$is_element) / nrow(pM.predictions.radius.strict.strat))

# Permissive set: set 8

fdr = region.cutoffs[["set8"]]$fdr

min.l2fc = region.cutoffs[["set8"]]$min.l2fc

min.baseMean = region.cutoffs[["set8"]]$min.baseMean

p1.elements = names(readRDS(paste0("../r_results/select_diff_regions/", 
                                   "p1_dep_nfia_ranges_with_sites_plus-strand_80pc_match-fdr_", fdr.matches, 
                                   "_in-regions-of_fdr", fdr, "_min-l2fc", min.l2fc, "_min-baseMean", min.baseMean, ".rds")))

p2.elements = names(readRDS(paste0("../r_results/select_diff_regions/", 
                                   "p2_dep_nfia_ranges_with_sites_plus-strand_80pc_match-fdr_", fdr.matches, 
                                   "_in-regions-of_fdr", fdr, "_min-l2fc", min.l2fc, "_min-baseMean", min.baseMean, ".rds")))

pM.elements = names(readRDS(paste0("../r_results/select_diff_regions/", 
                                   "pM_dep_nfia_ranges_with_sites_plus-strand_80pc_match-fdr_", fdr.matches, 
                                   "_in-regions-of_fdr", fdr, "_min-l2fc", min.l2fc, "_min-baseMean", min.baseMean, ".rds")))

p1.predictions.radius.permissive.strat = p1.predictions.radius.permissive %>% 
  rowwise() %>%
  mutate(is_element = (region_id %in% p1.elements)) %>%
  ungroup()

p2.predictions.radius.permissive.strat = p2.predictions.radius.permissive %>% 
  rowwise() %>%
  mutate(is_element = (region_id %in% p2.elements)) %>%
  ungroup()

pM.predictions.radius.permissive.strat = pM.predictions.radius.permissive %>% 
  rowwise() %>%
  mutate(is_element = (region_id %in% pM.elements)) %>%
  ungroup()

predictions.radius.permissive.strat = c(sum(p1.predictions.radius.permissive.strat$is_element) / nrow(p1.predictions.radius.permissive.strat),
                                        1 - sum(p1.predictions.radius.permissive.strat$is_element) / nrow(p1.predictions.radius.permissive.strat),
                                        sum(p2.predictions.radius.permissive.strat$is_element) / nrow(p2.predictions.radius.permissive.strat),
                                        1 - sum(p2.predictions.radius.permissive.strat$is_element) / nrow(p2.predictions.radius.permissive.strat),
                                        sum(pM.predictions.radius.permissive.strat$is_element) / nrow(pM.predictions.radius.permissive.strat),
                                        1 - sum(pM.predictions.radius.permissive.strat$is_element) / nrow(pM.predictions.radius.permissive.strat))

# TAD-based predictions

# Permissive set: set 8

fdr = region.cutoffs[["set8"]]$fdr

min.l2fc = region.cutoffs[["set8"]]$min.l2fc

min.baseMean = region.cutoffs[["set8"]]$min.baseMean

p1.elements = names(readRDS(paste0("../r_results/select_diff_regions/", 
                                   "p1_dep_nfia_ranges_with_sites_plus-strand_80pc_match-fdr_", fdr.matches, 
                                   "_in-regions-of_fdr", fdr, "_min-l2fc", min.l2fc, "_min-baseMean", min.baseMean, ".rds")))

p2.elements = names(readRDS(paste0("../r_results/select_diff_regions/", 
                                   "p2_dep_nfia_ranges_with_sites_plus-strand_80pc_match-fdr_", fdr.matches, 
                                   "_in-regions-of_fdr", fdr, "_min-l2fc", min.l2fc, "_min-baseMean", min.baseMean, ".rds")))

pM.elements = names(readRDS(paste0("../r_results/select_diff_regions/", 
                                   "pM_dep_nfia_ranges_with_sites_plus-strand_80pc_match-fdr_", fdr.matches, 
                                   "_in-regions-of_fdr", fdr, "_min-l2fc", min.l2fc, "_min-baseMean", min.baseMean, ".rds")))

p1.predictions.tad.permissive.strat = p1.predictions.tad.permissive %>% 
  rowwise() %>%
  mutate(is_element = (region_id %in% p1.elements)) %>%
  ungroup()

p2.predictions.tad.permissive.strat = p2.predictions.tad.permissive %>% 
  rowwise() %>%
  mutate(is_element = (region_id %in% p2.elements)) %>%
  ungroup()

pM.predictions.tad.permissive.strat = pM.predictions.tad.permissive %>% 
  rowwise() %>%
  mutate(is_element = (region_id %in% pM.elements)) %>%
  ungroup()

predictions.tad.permissive.strat = c(sum(p1.predictions.tad.permissive.strat$is_element) / nrow(p1.predictions.tad.permissive.strat),
                                     1 - sum(p1.predictions.tad.permissive.strat$is_element) / nrow(p1.predictions.tad.permissive.strat),
                                     sum(p2.predictions.tad.permissive.strat$is_element) / nrow(p2.predictions.tad.permissive.strat),
                                     1 - sum(p2.predictions.tad.permissive.strat$is_element) / nrow(p2.predictions.tad.permissive.strat),
                                     sum(pM.predictions.tad.permissive.strat$is_element) / nrow(pM.predictions.tad.permissive.strat),
                                     1 - sum(pM.predictions.tad.permissive.strat$is_element) / nrow(pM.predictions.tad.permissive.strat))

p = data.frame(category = rep(c("Element", "Other"), 3),
               domain = c(rep("p1", 2), rep("p2", 2), rep("pM", 2)),
               proportion = predictions.radius.strict.strat) %>%
  mutate(category = factor(category, levels = c("Element", "Other")),
         domain = factor(domain, levels = c("p1", "p2", "pM"))) %>%
  ggplot(aes(x = domain,
             y = proportion,
             fill = category)) +
    geom_col() +
    theme_classic()

ggsave(p, file = "../r_results/predict_correlated_deg_highest/plots/predictions_radius_strict_strat.pdf")

p = data.frame(category = rep(c("Element", "Other"), 3),
               domain = c(rep("p1", 2), rep("p2", 2), rep("pM", 2)),
               proportion = predictions.radius.permissive.strat) %>%
  mutate(category = factor(category, levels = c("Element", "Other")),
         domain = factor(domain, levels = c("p1", "p2", "pM"))) %>%
  ggplot(aes(x = domain,
             y = proportion,
             fill = category)) +
    geom_col() +
    theme_classic()

ggsave(p, file = "../r_results/predict_correlated_deg_highest/plots/predictions_radius_permissive_strat.pdf")

p = data.frame(category = rep(c("Element", "Other"), 3),
               domain = c(rep("p1", 2), rep("p2", 2), rep("pM", 2)),
               proportion = predictions.tad.permissive.strat) %>%
  mutate(category = factor(category, levels = c("Element", "Other")),
         domain = factor(domain, levels = c("p1", "p2", "pM"))) %>%
  ggplot(aes(x = domain,
             y = proportion,
             fill = category)) +
    geom_col() +
    theme_classic()

ggsave(p, file = "../r_results/predict_correlated_deg_highest/plots/predictions_tad_permissive_strat.pdf")
```

Count shared and unique NFIA-dependent region-gene pairs:

```{r, include=T}
predictions.radius.strict.strat.proportions.df = find_shared_features(paste0(p1.predictions.radius.strict.strat$region_id, "|",
                                                                             p1.predictions.radius.strict.strat$gene_name), 
                                                                      paste0(p2.predictions.radius.strict.strat$region_id, "|",
                                                                             p2.predictions.radius.strict.strat$gene_name),
                                                                      paste0(pM.predictions.radius.strict.strat$region_id, "|",
                                                                             pM.predictions.radius.strict.strat$gene_name),
                     plot.filename = "../r_results/predict_correlated_deg_highest/plots/predictions_radius_strict_strat_shared_and_unique_pairs.pdf")

predictions.radius.permissive.strat.proportions.df = find_shared_features(paste0(p1.predictions.radius.permissive.strat$region_id, "|",
                                                                                 p1.predictions.radius.permissive.strat$gene_name), 
                                                                          paste0(p2.predictions.radius.permissive.strat$region_id, "|",
                                                                                 p2.predictions.radius.permissive.strat$gene_name),
                                                                          paste0(pM.predictions.radius.permissive.strat$region_id, "|",
                                                                                 pM.predictions.radius.permissive.strat$gene_name),
                     plot.filename = "../r_results/predict_correlated_deg_highest/plots/predictions_radius_permissive_strat_shared_and_unique_pairs.pdf")

predictions.tad.permissive.strat.proportions.df = find_shared_features(paste0(p1.predictions.tad.permissive.strat$region_id, "|",
                                                                              p1.predictions.tad.permissive.strat$gene_name), 
                                                                       paste0(p2.predictions.tad.permissive.strat$region_id, "|",
                                                                              p2.predictions.tad.permissive.strat$gene_name),
                                                                       paste0(pM.predictions.tad.permissive.strat$region_id, "|",
                                                                              pM.predictions.tad.permissive.strat$gene_name),
                     plot.filename = "../r_results/predict_correlated_deg_highest/plots/predictions_tad_permissive_strat_shared_and_unique_pairs.pdf")
```

